---
title: "Juvenile snowshoe hares - exploratory report"
author: "By: Lauren Skube"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(effsize)
library(broom)
library(kableExtra)
library(lubridate)
```

### 1. Overview

Write an overview of the report here. 

```{r, include=FALSE, message=FALSE}
bonanza_hares <- read.csv(here("data" , "bonanza_hares.csv")) # Read in the data using the here() function
```


### 2. Data and analyses 

### 3. Exploratory findings 


#### A. Annual juvenile hare trap counts 

```{r, include=FALSE}
juv_hares <- bonanza_hares %>% 
  filter(age == "j") %>% # Create a subset that only contains juveniles
  mutate(date = mdy(date)) %>% # Change the date column to "date" class
  mutate(year = year(date)) # add a new column for the year 

juv_hares_counts <- juv_hares %>% 
  count(year) # Create a table containing the number of juvenile hares per year

juv_stats <- juv_hares_counts %>% 
  summarize(mean_counts = mean(n), 
            median_counts = median(n), 
            min_counts = min(n), 
            max_counts = max(n))

```

```{r juv_trappings_graph, echo=FALSE}
ggplot(data = juv_hares_counts, aes(x= year, y = n)) +
  geom_col(color = "darkslategrey", fill = "darkslategrey") +
  labs(x = "Year" , 
       y = "Total counts of juvenile trappings") +
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012)) +
  theme_minimal()
```

**Figure 1.** Total number of juvenile snowshoe hare trappings by year (1999 to 2012) across all grid sites in Bonanza Creek Experimental Forest, Alaska. Data: Kielland et al. (2017). 

Talk about major takeaways from the figure here. Talk about major takeaways from the figure here. Talk about major takeaways from the figure here. Talk about major takeaways from the figure here. 

#### B. Visualization of juvenile hare weights 
