---
title: "Juvenile snowshoe hares - exploratory report"
author: "By: Lauren Skube"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(effsize)
library(broom)
library(ggbeeswarm)
library(kableExtra)
library(lubridate)
```

### 1. Overview

This report provides an exploratory overview of differences between male and female juvenile snowshoe hares observed in Alaska's Bonanza Experimental Forest from 1999 to 2012 (Kielland et al. 2017). Juvenile weights are first compared by both trapping grid site and sex, and then by sex only. Lastly, the relationship between hind foot length and weight is explored. 

```{r, include=FALSE, message=FALSE}

 # Read in the data using the here() function
bonanza_hares <- read.csv(here("data" , "bonanza_hares.csv"))

```


### 2. Data and analyses 

Juvenile snowshoe hare observations and measurements were collected and made available by Dr. Knut Kielland and colleagues at the Bonanza Creek Long Term Ecological Research (LTER) site in Fairbanks, Alaska. The Bonanza Creek Experimental Forest is a 50km^2^ experimental forest located within the Tanana Valley State Forest. The data contain observations for 3197 snowshoe hares (including 378 juveniles) from 1999 to 2012 across three trapping grids: Bonanza Riparian (a flood zone along the Tanana River), Bonanza Mature (a mature black spruce forest 1 km inland), and Bonanza Black Spruce (a mature lowland forest area). Annual juvenile snowshoe hare trappings are counted for each year of the study, and juvenile weights are compared by sex and trapping grid. Following exploratory data visualization, juvenile weights between male and female hares are compared using a two-sample t-test ($\alpha$ = 0.05), and the differences between groups are described by Cohen's *d* effect size. The relationship between hind foot length and weight is explored by simple linear regression using Pearson's r as a descriptor of the degree of linear association between the two varibales. 

### 3. Exploratory findings 

#### A. Annual juvenile hare trap counts 

```{r, include=FALSE}
juv_hares <- bonanza_hares %>% 
  filter(age == "j") %>% # Create a subset that only contains juveniles
  mutate(date = mdy(date)) %>% # Change the date column to "date" class
  mutate(year = year(date)) %>% # Add a new column for the year 
  mutate(sex_full = case_when( # Add a new column with the full name of sex
    sex %in% c("f") ~ "Female" ,
    sex %in% c("m") ~ "Male"))
  
juv_grid_full <- juv_hares %>% # Create a subset with the full name for grid sites
  mutate(grid_full = case_when(
      grid %in% c("bonrip") ~ "Bonanza Riparian" ,
      grid %in% c("bonmat") ~ "Bonanza Mature" ,
      grid %in% c("bonbs") ~ "Bonzana Black Spruce"))

# Create a table containing the number of juvenile hare trappings per year
juv_hares_counts <- juv_hares %>% 
  count(year) 

# Summary statistics for juvenile trapping counts
juv_stats <- juv_hares_counts %>% 
  summarize(mean_counts = mean(n), 
            median_counts = median(n), 
            min_counts = min(n), 
            max_counts = max(n))
```

```{r juv_trappings_graph, echo=FALSE}

# Create a bar graph for juvenile trappings per year 
ggplot(data = juv_hares_counts, aes(x= year, y = n)) +
  geom_col(color = "darkslategrey", fill = "darkslategrey") +
  labs(x = "Year" , 
       y = "Total counts of juvenile trappings") +
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012)) + # Customize x scale
  theme_minimal()
```

**Figure 1.** Total number of juvenile snowshoe hare trappings by year (1999 to 2012) across all grid sites in the Bonanza Creek Experimental Forest, Alaska. Data: Kielland et al. (2017). 

Over the course of the study, the minimum count of juvenile hare trappings that occurred in a single year was 2 hares (2010). The maximum count of juvenile trappings that occurred in a single year was 126 hares (1999). The annual mean and median counts of juvenile hares were 31 and 18, respectively. Overall, the number of juvenile trappings tended to decrease over time, aside from two substantial increases relative to previous years, which occurred in 2005 and 2008. In future analyses, the juvenile hare population data could be standardized by converting observations to z-scores. This would allow annual juvenile trappings to be compared based on the number of standard deviations they are from the mean value.

#### B. Visualization of juvenile hare weights 

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Create a beeswarm plot for each sex separated by grid
ggplot(data = juv_grid_full, aes(x = sex_full, y = weight)) +
  geom_beeswarm(aes(color = sex), 
                size = 1.5,
                cex = 2, 
                alpha = 0.8) + 
  facet_wrap(~grid_full) +
  geom_boxplot(fill=NA, # add a boxplot on top of beeswarm
               width = 0.6, 
               outlier.color = "black") + # Show outliers as black dots so they stand out 
  theme_light() +
  theme(legend.position = "none") + # remove the legend 
  labs(x = "Sex", 
       y = "Weight (g)")
```
**Figure 2.** Weight observations (g) of juvenile male and female snowshoe hares by trapping grid site (Bonanza Mature forest, Bonanza Riparian forest, and Bonanza Black Spruce forest). Pink (female) and teal (male) points indicate individual observations for weight (grams) of juvenile snowshoe hares. Box endpoints indicate the 25th and 75th percentile values; the black line indicates the median for each sex at the different grid sites. Outliers are indicated by black points. Data: Kielland et al. (2017).

Juvenile snowshoe hare weights were compared by sex and grid site. There were notable outliers in the female sample from the Bonanza Riparian site and in the male sample from the Bonanza Black Spruce site. Median values for each sex were similar across the different grid sites.  Each group had a relatively large spread, with many overlapping points between groups of both males and females and across grid sites. Some of the juvenile groups, including Bonanza Riparian females and Bonanza Mature males, appeared to have a bimodal distribution rather than relative distributions. 

#### **C. Juvenile weight comparison (male and female snowshoe hares)**

**Table 1.** Descriptive statistics (mean, standard deviation, and sample size) for weight of female and male juvenile snowshoe hares. 

```{r, echo=FALSE, message=FALSE}

# Create a table of summary statistics for juvenile hare weights
juv_hares %>% 
  drop_na(weight) %>% 
  drop_na(sex_full) %>% # Remove NA values 
  group_by(sex_full) %>% 
  summarise(
    mean_weight = mean(weight) ,
    sd_weight = sd(weight),
    sample_size = n()
    ) %>% 
  # format summary statistics in a nice table
  kbl(col.names = c("Sex" , "Mean weight (g)", "Standard deviation" , "Sample size")) %>% 
  kable_styling()


```

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Create a subset that does not include NA values 
juv_hares_clean <- juv_hares %>% 
  drop_na(weight)

# Actual difference in means between males and females: 90.568 g 
945.8589-855.3909

# Percent difference: 
90.468/((945.8589+855.3909)/2) *100

# T-Test code:

# Create a vector of weight observations for males
m_weight <- juv_hares_clean %>% 
  filter(sex == "m") %>% 
  pull(weight) 

# Create a vector of weight observations for females 
f_weight <- juv_hares_clean %>% 
  filter(sex == "f") %>% 
  pull(weight) 

# Run the t-test using the vectors
weight_ttest <- t.test(m_weight, f_weight) 

weight_ttest_tidy <- tidy(weight_ttest)

# Get the means/sd for both groups 
male_mean <- mean(m_weight)
female_mean <- mean(f_weight)
male_sd <- sd(m_weight)
female_sd <- sd(f_weight)

# Effect size (Cohen's d): 
weight_effsize <- cohen.d(m_weight , f_weight)


```

```{r, echo=FALSE, include=FALSE}
# Exploratory graphs to check assumptions of t-test

# Create a quantile-quantile graph to compare sample to normality
ggplot(data = juv_hares_clean, aes(sample = weight)) + 
  geom_qq() +
  facet_wrap(~sex)

# Create a histogram to compare sample to normality 
ggplot(data = juv_hares_clean, aes(x = weight)) + 
  geom_histogram() + 
  facet_wrap(~sex)


```

Distributions of weights for male and female juvenile snowshoe hares were relatively normal and symmetric. On average, male juvenile snowshoe hares weighed more than juvenile females (`r round(male_mean,2)` $\pm$ `r round(male_sd, 2)` and `r round(female_mean,2)` $\pm$ `r round(female_sd, 2)` g, respectively; mean $\pm$ 1 standard deviation). The absolute difference in means is `r round(male_mean, 2) - round(female_mean, 2)` g (a `r round((male_mean - female_mean)/((male_mean + female_mean)/2)*100, 2)`% difference), and is significant (Welch's two-sample t-test: t(`r round(weight_ttest_tidy$parameter,2)`) = `r round(weight_ttest_tidy$statistic,2)`, p = `r round(weight_ttest_tidy$p.value,2)`. While the difference in means is significant, the effect size is small (Cohen's *d* = (`r round(weight_effsize$estimate,2)`). 


#### D. Relationship between juvenile hare weight and hind foot length 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create a scatterplot for weight vs hind foot length
ggplot(data = juv_hares_clean, aes(x = hindft, y=weight)) + 
  geom_point(color = "darkslategrey") +
  labs(x = "Hind foot length (mm)" , 
       y = "Weight (g)") +
  geom_smooth(method = "lm", se = FALSE, color = "gray30") +
  theme_light()
```

**Figure 3.** Relationship between hind foot length (mm) and weight (g) for juvenile snowshoe hares across all grid sites. Points indicate individual juvenile hare measurements. Kielland et al. (2017).


#### C. Linear regression analysis 
```{r, echo=FALSE, include=FALSE, message=FALSE}
# Try out linear regression:

# model it with the lm() function 
juveniles_lm <- lm(weight ~ hindft, data = juv_hares_clean)

# Get tidy version of model output
juvenile_lm_tidy <- tidy(juveniles_lm)
juvenile_lm_glance <- glance(juveniles_lm)

# Use Pearson's r to test association between variables 
juvenile_cor <- cor.test(juv_hares_clean$weight, juv_hares_clean$hindft)

juvenile_cor_tidy <- tidy(juvenile_cor)
# Pearson's r = 0.55
```


```{r, include=FALSE, echo=FALSE}

# Check residual assumptions 
plot(juveniles_lm)

# first plot shows a clear trend in variance - assumption of homoscedasticity is violated, residuals are not distributed evenly or randomly
# second plot - shows residuals are not normally distributed about the model 
# fourth plot - labeled observations are outliers that may have too much leverage in the model 
```

Simple linear regression was used to explore the relationship between juvenile snowshoe hare hind foot length (mm) and weight (g) across all grid sites, including both male and female juveniles. A significant regression model was found (p <0.001, R^2^ = `r round(juvenile_lm_glance$r.squared,2)`), with an average slope of $\beta$ = `r round (juvenile_lm_tidy$estimate[2],2)` g/mm^-1^ (i.e., 30% of variance in juvenile snowshoe weight is explained by hind foot length, and for each one mm increase in hind foot length we expect an average increase in weight of `r round(juvenile_lm_tidy$estimate[2], 2)` g). Pearson's *r* is equal to `r round(juvenile_cor_tidy$estimate,2)`, indicating a moderate association between hind foot length and weight. However, diagnostic plots (not included) show a trend in residual variance, violating the assumption of homoscedasticity. Diagnostic plots also show that residuals are not normally distributed about the model and that there are outliers that may have too much leverage on the model. Due to these violations of assumptions, simple linear regression may not be accurate to describe the model. 

### Summary 

Exploratory data analysis reveals the following findings: 

- Counts of juvenile snowshoe hares have decreased dramatically since 1999. 
- On average, male juvenile snowshoe hares weigh more than juvenile females (the difference is significant, but the effect size is small). 
- Hind foot length appears linearly related to weight (the relationship is significant, with Pearson's r showing a moderate association between variables). However, diagnostic plots show that assumptions regarding residuals may be violated 
- The slope of the linear regression is 9.52, which is the expected average increase in weight (g) for every 1 mm increase in hind foot length. 

### Citations 

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14


