---
title: "Juvenile snowshoe hares - exploratory report"
author: "By: Lauren Skube"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(effsize)
library(broom)
library(ggbeeswarm)
library(kableExtra)
library(lubridate)
```

### 1. Overview

Write an overview of the report here. 

```{r, include=FALSE, message=FALSE}
bonanza_hares <- read.csv(here("data" , "bonanza_hares.csv")) # Read in the data using the here() function

```


### 2. Data and analyses 

### 3. Exploratory findings 

#### A. Annual juvenile hare trap counts 

```{r, include=FALSE}
juv_hares <- bonanza_hares %>% 
  filter(age == "j") %>% # Create a subset that only contains juveniles
  mutate(date = mdy(date)) %>% # Change the date column to "date" class
  mutate(year = year(date)) %>% # Add a new column for the year 
  mutate(sex_full = case_when( # Add a new column with the full name of the sex 
    sex %in% c("f") ~ "Female" ,
    sex %in% c("m") ~ "Male"))
  
juv_grid_full <- juv_hares %>% # Create a subset with the full name for grid sites
  mutate(grid_full = case_when(
      grid %in% c("bonrip") ~ "Bonanza Riparian" ,
      grid %in% c("bonmat") ~ "Bonanza Mature" ,
      grid %in% c("bonbs") ~ "Bonbs"))

juv_hares_counts <- juv_hares %>% 
  count(year) # Create a table containing the number of juvenile hare trappings per year

# Summary statistics for juvenile trapping counts
juv_stats <- juv_hares_counts %>% 
  summarize(mean_counts = mean(n), 
            median_counts = median(n), 
            min_counts = min(n), 
            max_counts = max(n))

```

```{r juv_trappings_graph, echo=FALSE}

# Create a bar graph for juvenile trappings per year 
ggplot(data = juv_hares_counts, aes(x= year, y = n)) +
  geom_col(color = "darkslategrey", fill = "darkslategrey") +
  labs(x = "Year" , 
       y = "Total counts of juvenile trappings") +
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012)) + # Customize x scale
  theme_minimal()
```

**Figure 1.** Total number of juvenile snowshoe hare trappings by year (1999 to 2012) across all grid sites in Bonanza Creek Experimental Forest, Alaska. Data: Kielland et al. (2017). 

Over the course of the study, the minimum count of juvenile hare trappings that occurred in a single year was 2 hares (2010). The maximum count of juvenile trappings that occurred in a single year was 126 hares (1999). The annual mean and median counts of juvenile hares were 31 and 18, respectively. Overall, the number of juvenile trappings tended to decrease over time, aside from two substantial increases relative to previous years, which occurred in 2005 and 2008. In future analyses, the juvenile hare population data could be standardized by converting observations to z-scores. This would allow annual juvenile trappings to be compared based on the number of standard deviations they are from the mean value.

#### B. Visualization of juvenile hare weights 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create a beeswarm plots for each sex separated by grid
ggplot(data = juv_grid_full, aes(x = sex_full, y = weight)) +
  geom_beeswarm(aes(color = sex), 
                size = 1.5,
                cex = 2, 
                alpha = 0.8) + 
  facet_wrap(~grid_full) +
  geom_boxplot(fill=NA, # add a boxplot on top of beeswarm
               width = 0.6, 
               outlier.color = "black") + 
  theme_light() +
  theme(legend.position = "none") +
  labs(x = "Sex", 
       y = "Weight (g)")
```

```{r, echo=FALSE, warning=FALSE, include=FALSE}

# Create boxplot with male and female weights by grid site
ggplot(juv_grid_full, aes(x= sex_full, y = weight, fill = sex_full)) +
  geom_boxplot(size = 0.75, # Increase thickness of lines
               outlier.color = "black") + # Show outliers in black 
  facet_wrap(~grid_full) + # Separate by grid site
  labs(x = "Sex", 
       y = "Weight (g)") + 
  #Overlay with a jitterplot to show individual observations
  geom_jitter(alpha = 0.4, # Customize transparency of points
              width = 0.15) + # Reduce jitter width
  theme_light() +
  theme(legend.position = "none") # Remove the legend

```
**Figure 2.** Weight observations (g) of juvenile male and female snowshoe hares for each trapping grid site (Bonanza Mature forest, Bonanza Riparian forest, and Bonbs). Pink (female) and teal (male) points indicate individual observations for weight (grams) of juvenile snowshoe hares. Box endpoints indicate the 25th and 75th percentile values; the black line indicates the median for each sex at the different grid sites. Outliers are indicated by black points. Data: Kielland et al. (2017).

Juvenile snowshoe hare weights were compared by sex and grid site. There were notable outliers in the female sample from the Bonanza Riparian site and in the male sample from the Bonbs site. Median values were similar across the different grids for both male and female groups. 
Each group had a relatively large spread, with many overlapping points between groups of both males and females and across grid sites. Some of the juvenile groups, including Bonanza Riparian females and Bonanza Mature males, appeared to have a bimodal distribution rather than relative distributions. 

#### **C. Juvenile weight comparison (male and female snowshoe hares)**

**Table 1.** Descriptive statistics (mean, standard deviation, and sample size) for weight of female and male juvenile snowshoe hares. 

```{r, echo=FALSE, message=FALSE}

# Create a table of summary statistics for juvenile hare weights
juv_hares %>% 
  drop_na(weight) %>% 
  drop_na(sex_full) %>% # Remove NA values 
  group_by(sex_full) %>% 
  summarise(
    mean_weight = mean(weight) ,
    sd_weight = sd(weight),
    sample_size = n()
    ) %>% 
  # format summary statistics in a nice table
  kbl(col.names = c("Sex" , "Mean weight (g)", "Standard deviation" , "Sample size")) %>% 
  kable_styling()


```
```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Create a subset that does not include NA values 
juv_hares_clean <- juv_hares %>% 
  drop_na(weight)

# Actual difference in means between males and females: 90.568 g 
945.8589-855.3909

# Percent difference: 
90.468/((945.8589+855.3909)/2) *100

# T-Test code:

# Create a vector of weight observations for males
m_weight <- juv_hares_clean %>% 
  filter(sex == "m") %>% 
  pull(weight) 

# Create a vector of weight observations for females 
f_weight <- juv_hares_clean %>% 
  filter(sex == "f") %>% 
  pull(weight) 

# Run the t-test 
weight_ttest <- t.test(m_weight, f_weight) 

# Effect size (Cohen's d): 
cohen.d(m_weight , f_weight)



weight_ttest

```

```{r, echo=FALSE, include=FALSE}
# Exploratory graphs to check assumptions of t-test

# Create a quantile-quantile graph to compare sample to normality
ggplot(data = juv_hares_clean, aes(sample = weight)) + 
  geom_qq() +
  facet_wrap(~sex)

# Create a histogram to compare sample to normality 
ggplot(data = juv_hares_clean, aes(x = weight)) + 
  geom_histogram() + 
  facet_wrap(~sex)


```

Distributions of weights for male and female juvenile snowshoe hares were relatively normal and symmetric. On average, male juvenile snowshoe hares weigh more than juvenile females. Juvenile males have a mean weight of 945.86 grams (sd = 333.22g), and females have a mean weight of 855.39 grams (sd = 292.25g). The absolute difference in means is 90.47 g (10.05% difference), and is significant (Welch's two-sample t-test: t(325.02) = 2.71, p = 0.007). While the difference in means is significant, the effect size is small (Cohen's *d* = 0.29). 


#### D. Relationship between juvenile hare weight and hind foot length 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create a scatterplot for weight vs hind foot length
ggplot(data = juv_hares_clean, aes(x = hindft, y=weight)) + 
  geom_point(color = "darkslategrey") +
  labs(x = "Hind foot length (mm)" , 
       y = "Weight (g)") +
  geom_smooth(method = "lm", se = FALSE, color = "gray30") +
  theme_light()
```

**Figure 3.** Relationship between hind foot length (mm) and weight (g) for all juvenile snowshoe hares. Points indicate individual juvenile hare measurements. Kielland et al. (2017).


#### C. Linear regression analysis 
```{r, echo=FALSE, include=FALSE, message=FALSE}
# Try out linear regression:

# model it with the lm() function 
juveniles_lm <- lm(weight ~ hindft, data = juv_hares_clean)

# Get tidy version of model output
juvenile_lm_tidy <- tidy(juveniles_lm)
juvenile_lm_glance <- glance(juveniles_lm)

juvenile_cor <- cor.test(juv_hares_clean$weight, juv_hares_clean$hindft)

juvenile_cor_tidy <- tidy(juvenile_cor)
# Peason's r = 0.55
```


```{r, include=FALSE, echo=FALSE}

# Check residual assumptions 
plot(juveniles_lm)

# first plot shows a clear trend in variance - assumption of homoscedasticity is violated, residuals are not distributed evenly or randomly
# second plot - shows residuals are not normally distributed about the model 
# fourth plot - labeled observations are outliers that may have too much leverage in the model 
```

Simple linear regression was used to explore the relationship between juvenile snowshoe hare hind foot length (mm) and weight (g) across all grid sites, including both male and female juveniles. A significant regression model was found (β = 9.52, F(1,247) = 105.2, p < 0.001) with an R^2^ of 0.299 (i.e., 29.9% of variance in juvenile snowshoe weight is explained by hind foot length, and for each one mm increase in hind foot length we expect an average increase in weight of 9.52 grams). Pearson's *r* is equal to 0.55, indicating a moderate association between hind foot length and weight. However, diagnostic plots (not included) show a trend in variance, violating the assumption of homoscedasticity. Diagnostic plots also show that residuals are not normally distributed about the model and that there are outliers that may have too much leverage on the model. Due to these violations of assumptions, linear regression may not be describe the model well. 
